SUPPLEMENTARY MATERIALS for\\ \strut \\ \textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\ Beyond repeated measures ANOVA and Linear Mixed Models\\ \strut \\ APPENDIX A: WORKFLOW FOR GAMs FOR BIOMEDICAL LONGITUDINAL DATA

### 2021-12-23 Manuscripts/Manuscript_by_chapters-SIM_Revisions_final/appendices/Appendix_A.tex
New directory, knit separate appendices, separate directory for scripts

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
%\usepackage{lineno}
\usepackage[affil-it,blocks]{authblk}
\usepackage{hyperref}
\usepackage{graphicx}
%\usepackage[nomarkers,figuresonly]{endfloat}
%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{latex_docs/orcid}}

@@

%to have page numbers with letters in the Appendix:
% %solution from https://tex.stackexchange.com/questions/59572/custom-page-numbering-for-appendix
\newcommand{\appendixpagenumbering}{
  \break
  \pagenumbering{arabic}
  \renewcommand{\thepage}{\thesection-\arabic{page}}
}

@@

%authors
\author{Ariel I. Mundo \orcidaffila{}}
\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
\author{John R. Tipton \orcidaffilb{}}
\affil{Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
\author{Timothy J. Muldoon*}
\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
\affil{tmuldoon@uark.edu}

@@

%\lstset{
%  breaklines=true,
%  stringstyle=\ttfamily,
%  backgroundcolor=\color{gray}
%}
\usepackage{placeins}
\usepackage{subfig}
\usepackage{breqn}
\usepackage[font={small}]{caption}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
Beyond repeated measures ANOVA and Linear Mixed Models\\
\strut \\
SUPPLEMENTARY MATERIALS: APPENDIX}
\date{\vspace{-2.5em}}

@@

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the \emph{repeated measures analysis of variance} (rm-ANOVA) or more recently, \emph{linear mixed models} (LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.

In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We present the basic theory of GAMs, and using reported trends of oxygen saturation in tumors we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R. We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing). To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.

\hypertarget{keywords}{%
\section*{Keywords}\label{keywords}}
\addcontentsline{toc}{section}{Keywords}

cancer biology; tumor response; generalized additive models; simulation; R

\hypertarget{background}{%
\section{Background}\label{background}}

Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study). Biomedical research frequently uses longitudinal studies to analyze the evolution of a ``treatment'' effect across multiple time points, with subjects of analysis ranging from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others. Tumor response {[}\protect\hyperlink{ref-roblyer2011}{1}--\protect\hyperlink{ref-demidov2018}{4}{]}, antibody expression {[}\protect\hyperlink{ref-ritter2001}{5},\protect\hyperlink{ref-roth2017}{6}{]}, and cell metabolism {[}\protect\hyperlink{ref-jones2018}{7},\protect\hyperlink{ref-skala2010}{8}{]} are examples of different situations where researchers have used longitudinal designs to study some physiological response. Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals{[}\protect\hyperlink{ref-greening2018}{9}{]}, to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients {[}\protect\hyperlink{ref-sio2016}{10}{]}, to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients {[}\protect\hyperlink{ref-kamstra2015}{11}{]}.

Traditionally, a ``frequentist'' or ``classical'' statistical paradigm is used in biomedical research to derive inferences from a longitudinal study. The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times {[}\protect\hyperlink{ref-wagenmakers2008}{12}{]}, and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an \emph{analysis of variance over repeated measures} (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model makes three assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as \emph{complete observations}) {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-schober2018}{14}{]}.

The expected linear behavior of the response through time is a key requisite in rm-ANOVA {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}. This ``linearity assumption'' in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In longitudinal biomedical research, non-linear trends are the norm rather than the exception. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings {[}\protect\hyperlink{ref-roblyer2011}{1},\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.

A \emph{post hoc} analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a \emph{p-value}, which in turn is used as a measure of significance.
Although it is possible that a \emph{post hoc} analysis of rm-ANOVA is able to find ``significant'' \emph{p-values}( \emph{p}\textless0.05) from data with non-linear trends, the validity of such a metric is dependent on how adequate the model fits the data. In other words, \emph{p-values} are valid only if the model and the data have good agreement; if that is not the case, a ``Type III'' error (known as ``model misspecification'') occurs{[}\protect\hyperlink{ref-dennis2019}{17}{]}. For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting \emph{p-values} and parameter estimates to be invalid {[}\protect\hyperlink{ref-wang2019}{18}{]}.

Additionally, the \emph{p-value} itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or \(\alpha\)) {[}\protect\hyperlink{ref-liu2010}{19},\protect\hyperlink{ref-halsey2015}{20}{]}, consequently biasing the conclusions of the study. Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni {[}\protect\hyperlink{ref-abdi2010}{21}{]}), but they in turn reduce statistical power (1-\(\beta\)){[}\protect\hyperlink{ref-nakagawa2004}{22}{]}, and lead to increased Type II error (failing to reject the null hypothesis when it is false) {[}\protect\hyperlink{ref-gelman2012}{23},\protect\hyperlink{ref-albers2019}{24}{]}. Therefore, the tradeoff of \emph{post hoc} comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.

On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the \emph{compound symmetry assumption}) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases {[}\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}. Corrections can be made in rm-ANOVA in the absence of compound symmetry {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]}, but the effectiveness of the correction is limited by the size of the sample, the number of measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and group sizes {[}\protect\hyperlink{ref-keselman2001}{29}{]}. In the case of biomedical research, where living subjects are frequently used, sample sizes are often not ``large'' due to ethical and budgetary reasons, {[}\protect\hyperlink{ref-charan2013}{30}{]} which might cause the corrections for lack of compound symmetry to be ineffective.

Due to a variety of causes, the number of observations during a study can be different between all subjects. For example, in a clinical trial patients may voluntarily withdraw, whereas attrition due to injury or weight loss in preclinical animal studies is possible. It is even plausible that unexpected complications with equipment or supplies arise that prevent the researcher from collecting measurements at certain time points. In each of these missing data scenarios, the \emph{complete observations} assumption of classical rm-ANOVA is violated. When incomplete observations occur, a rm-ANOVA model is fit by excluding all subjects with missing observations from the analysis {[}\protect\hyperlink{ref-gueorguieva2004}{13}{]}. This elimination of partially missing data from the analysis can result in increased costs if the desired statistical power is not met with the remaining observations, because it would be necessary to enroll more subjects. At the same time, if the excluded observations contain insightful information that is not used, their elimination from the analysis may limit the demonstration of significant differences between groups.

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data. The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data {[}\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}{]}. Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}. However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}{]}. And even more importantly, LMEMs also assume by default a linear relationship between the response and time {[}\protect\hyperlink{ref-pinheiro2006}{15}{]} (polynomial effects can be used, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .

As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the trend of the response, there is a need for biomedical researchers to explore the use of statistical tools that allow the data (and not a model assumption) to determine the trend of the fitted model and to enable appropriate inference. In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data. Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data. Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data {[}\protect\hyperlink{ref-rose2012}{32}--\protect\hyperlink{ref-simpson2018}{34}{]}, health-environment interactions {[}\protect\hyperlink{ref-yang2012}{35}{]} and the dynamics of government in political science {[}\protect\hyperlink{ref-beck1998}{36}{]}. There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements {[}\protect\hyperlink{ref-wood2017}{37}{]}, and 3) can easily accommodate incomplete observations. Therefore, GAMs provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.

The current advances in programming languages designed for statistical analysis (specifically R), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs. In particular, R {[}\protect\hyperlink{ref-r}{38}{]} has an extensive collection of documentation and functions to fit GAMs in the package \emph{mgcv} {[}\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2016}{39}{]} that speed up the initial stages of the analysis and enable the use of advanced modeling structures (e.g.~hierarchical models, confidence interval comparisons) without requiring advanced programming skills. At the same time, R has many tools that simplify data simulation.

Data simulation methods are an emerging technique that allow the researcher to create and explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation. In addition, simulation can be also used for power calculations and study design questions {[}\protect\hyperlink{ref-haverkamp2017}{28}{]}.

This work provides biomedical researchers with a clear understanding of the theory and the practice of using GAMs to analyze longitudinal data using by focusing on four areas. First, the limitations of LMEMs and rm-ANOVA regarding an expected trend of the response, constant correlation structures, and missing observations are explained in detail. Second, the key theoretical elements of GAMs are presented using clear and simple mathematical notation while explaining the context and interpretation of the equations. Third, we illustrate the type of non-linear longitudinal data that often occurs in biomedical research using simulated data that reproduces patterns in previously reported studies {[}\protect\hyperlink{ref-vishwanath2009}{16}{]}. The simulated data experiments highlight the differences in inference between rm-ANOVA, LMEMs and GAMs on data similar to what is commonly observed in biomedical studies. Finally, reproducibility is emphasized by providing the code to generate the simulated data and the implementation of different models in R, in conjunction with a step-by-step guide demonstrating how to fit models of increasing complexity.

In summary, this work will allow biomedical researchers to identify when the use of GAMs instead of rm-ANOVA or LMEMs is appropriate to analyze longitudinal data, and provide guidance on the implementation of these models to improve the standards for reproducibility in biomedical research.

\hypertarget{challenges-presented-by-longitudinal-studies}{%
\section{Challenges presented by longitudinal studies}\label{challenges-presented-by-longitudinal-studies}}

\hypertarget{the-repeated-measures-anova-and-linear-mixed-model}{%
\subsection{The repeated measures ANOVA and Linear Mixed Model}\label{the-repeated-measures-anova-and-linear-mixed-model}}

The \emph{repeated measures analysis of variance} (rm-ANOVA) and the \emph{linear mixed model} (LMEM) are the most commonly used statistical analysis for longitudinal data in biomedical research. These statistical methodologies require certain assumptions for the model to be valid. From a practical view, the assumptions can be divided in three areas: 1) an assumed relationship between covariates and response, 2) a constant correlation between measurements, and, 3) complete observations for all subjects. Each one of these assumptions is discussed below.

\hypertarget{assumed-relationship}{%
\subsection{Assumed relationship}\label{assumed-relationship}}

\hypertarget{the-repeated-measures-anova-case}{%
\subsubsection{The repeated measures ANOVA case}\label{the-repeated-measures-anova-case}}

In a longitudinal biomedical study, two or more groups of subjects (e.g., human subject, mice, samples) are subject to different treatments (e.g., a ``treatment'' group receives a novel drug or intervention vs.~a ``control'' group that receives a placebo), and measurements from each subject within each group are collected at specific time points. The collected response is modeled with \emph{fixed} components. The \emph{fixed} component can be understood as a constant value in the response which the researcher is interested in measuring, i.e., the average effect of the novel drug/intervention in the ``treatment'' group.

Mathematically speaking, a rm-ANOVA model with an interaction can be written as

\begin{equation}
y_{ijt} = \beta_0+\beta_1 \times treatment_{j} +\beta_2 \times time_{t} +\beta_3 \times time_{t}\times treatment_{j}+\varepsilon_{ijt},\\ 
\label{eq:linear-model}
\end{equation}

In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of treatment (\(treatment_j\)), time (\(time_t\)), and their interaction \(time_t\times treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively. Independent errors \(\varepsilon_{ijt}\) represent random variation from the sampling process assumed to be \(\stackrel{i.i.d.}\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug'', or ``saline'' vs.~``chemotherapy''). Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B). With this notation, the linear model then can be expressed as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_2\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
\beta_0 + \beta_1+\beta_2 \times time_{t} +\beta_3 \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
\end{cases}
\label{eq:ANOVA-by-group}
\end{equation}

To further simplify the expression, substitute \(\widetilde{\beta_{0}}=\beta_0+\beta_{1}\) and \(\widetilde{\beta_{1}}=\beta_{2}+\beta_{3}\) in the equation for Group B. This substitution allows for a different intercept and slope for Groups A and B. The model is then written as

\begin{equation}
y_{ijt} = \begin{cases}
\beta_0 + \beta_2\times time_{t}+\varepsilon_{ijt}   & \mbox{if Group A},\\
\widetilde{\beta_{0}} + \widetilde{\beta_1} \times time_{t}+\varepsilon_{ijt}  & \mbox{if Group B}.\\
\end{cases}
\label{eq:ANOVA-lines}
\end{equation}

Presenting the model in this manner makes clear that when treating different groups, an rm-ANOVA model is able to accommodate non-parallel lines in each case (different intercepts and slopes per group). In other words, the rm-ANOVA model ``expects'' a linear relationship between the covariates and the response. This means that either presented as Equations \eqref{eq:linear-model}, \eqref{eq:ANOVA-by-group} or \eqref{eq:ANOVA-lines}, an rm-ANOVA model is only able to accommodate linear patterns in the data. If the data show non-linear trends, the rm-ANOVA model will approximate this behavior with non-parallel lines.

\hypertarget{LMEM-case}{%
\subsubsection{The Linear Mixed Model (LMEM) Case}\label{LMEM-case}}

A LMEM is a class of statistical models that incorporates \emph{fixed effects} to model the relationship between the covariates and the response, and \emph{random effects} to model subject variability that is not the primary focus of the study but that might be important to account for{[}\protect\hyperlink{ref-pinheiro2006}{15},\protect\hyperlink{ref-west2014}{40}{]}. A LMEM with interaction between time and treatment for a longitudinal study can be written as

\begin{equation}
y_{ijt} = \beta_0+ \beta_1 \times treatment_{j}+ \beta_2 \times time_{t} + \beta_3 \times time_{t}\times treatment_{j}+\alpha_{ij} +\varepsilon_{ijt}\\.
\label{eq:LMEM}
\end{equation}

When Equations \eqref{eq:linear-model} and \eqref{eq:LMEM} are compared, it is noticeable that LMEMs and rm-ANOVA have the same construction regarding the \emph{fixed effects} of time and treatment, but that the LMEM incorporates an additional source of variation (the term \(\alpha_{ij}\)). This term \(\alpha_{ij}\) corresponds to the \emph{random effect}, accounting for variability in each subject (subject\(_i\)) within each group (group\(_j\)). The \emph{random} component can also be understood as modeling some ``noise'' in the response, but that does not arise from the sampling error term \(\varepsilon_{ijt}\) from Equations \eqref{eq:linear-model} through \eqref{eq:ANOVA-lines}.

For example, if the blood concentration of a drug is measured in certain subjects in the early hours of the morning while other subjects are measured in the afternoon, it is possible that the difference in the collection time introduces some ``noise'' in the data that needs to be accounted for. As the name suggests, this ``random'' variability needs to be modeled as a variable rather than as a constant value. The random effect \(\alpha_{ij}\) in Equation \eqref{eq:LMEM} is assumed to be \(\alpha_{ij} \sim N(0,\sigma^2_\alpha)\). In essence, the \emph{random effect} in a LMEM enables fitting models with different intercepts at the subject-level{[}\protect\hyperlink{ref-pinheiro2006}{15}{]}. However, the expected linear relationship of the covariates and the response in Equation \eqref{eq:linear-model} and in Equation \eqref{eq:LMEM} is essentially the same, representing a major limitation of LMEMs to fit a non-linear response.

Of note, LMEMs are capable of fitting non-linear trends using an ``empirical'' approach (using polynomial fixed effects instead of linear effects such as in Equation \eqref{eq:LMEM}), which is described in detail by Pinheiro and Bates {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}. However, polynomial fits have limited predictive power, cause bias on the boundaries of the covariates {[}\protect\hyperlink{ref-beck1998}{36}{]}, and more importantly, their lack of biological or mechanistic interpretation limits their use in biomedical studies {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}.

\hypertarget{covariance-in-rm-anova-and-lmems}{%
\subsection{Covariance in rm-ANOVA and LMEMs}\label{covariance-in-rm-anova-and-lmems}}

In a longitudinal study there is an expected \emph{covariance} between repeated measurements on the same subject, and because repeated measures occur in the subjects within each group, there is a \emph{covariance} between measurements at each time point within each group. The \emph{covariance matrix} (also known as the variance-covariance matrix) is a matrix that captures the variation between and within subjects in a longitudinal study{[}\protect\hyperlink{ref-wolfinger1996}{41}{]} (For an in-depth analysis of the covariance matrix see West{[}\protect\hyperlink{ref-west2014}{40}{]} and Weiss{[}\protect\hyperlink{ref-weiss2005}{42}{]}).

In the case of an rm-ANOVA analysis, it is typically assumed that the covariance matrix has a specific construction known as \emph{compound symmetry} (also known as ``sphericity'' or ``circularity''). Under this assumption, the between-subject variance and within-subject correlation are constant across time {[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-weiss2005}{42},\protect\hyperlink{ref-geisser1958}{43}{]}. However, it has been shown that this condition is frequently not justified because the correlation between measurements tends to change over time {[}\protect\hyperlink{ref-maxwell2017}{44}{]}; and is higher between consecutive measurements {[}\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-ugrinowitsch2004}{25}{]}. Although corrections can be made (such as Huyhn-Feldt or Greenhouse-Geisser){[}\protect\hyperlink{ref-huynh1976}{26},\protect\hyperlink{ref-greenhouse1959}{27}{]} their effectiveness is dependent on sample size and number of repeated measurements{[}\protect\hyperlink{ref-haverkamp2017}{28}{]}, and it has been shown that corrections are not robust if the group sizes are unbalanced {[}\protect\hyperlink{ref-keselman2001}{29}{]}. Because biomedical longitudinal studies are often limited in sample size and can have an imbalanced design, the corrections required to use an rm-ANOVA model may not be able to provide a reasonable adjustment that makes the model valid.

In the case of LMEMs, one key advantage over rm-ANOVA is that they allow different structures for the variance-covariance matrix including exponential, autoregressive of order 1, rational quadratic and others {[}\protect\hyperlink{ref-pinheiro2006}{15}{]}. Nevertheless, the analysis required to determine an appropriate variance-covariance structure for the data can be a challenging process by itself. Overall, the spherical assumption for rm-ANOVA may not capture the natural variations of the correlation in the data, and can bias the inferences from the analysis.

\hypertarget{unbalanced-data}{%
\subsection{Unbalanced data}\label{unbalanced-data}}

In a longitudinal study, it is frequently the case that the number of observations is different across subjects. In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as ``missing'', ``incomplete'', or (more generally speaking) unbalanced data {[}\protect\hyperlink{ref-molenberghs2004}{45}{]}. The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis. This is inconvenient because the remaining subjects might not accurately represent the population and statistical power is affected by this reduction in sample size {[}\protect\hyperlink{ref-ma2012}{46}{]}.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
%\usepackage{lineno}
\usepackage[affil-it,blocks]{authblk}
\usepackage{hyperref}
\usepackage{graphicx}
%\usepackage[nomarkers,figuresonly]{endfloat}
%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{latex_docs/orcid}}

@@

%start sections and page numbers with A
\setcounter{page}{1}
\renewcommand\thesection{A}
\renewcommand\thesubsection{\thesection.\arabic{subsection}}
\renewcommand{\thepage}{A-\arabic{page}}

@@

%authors
\author{Ariel I. Mundo \orcidaffila{}}
%\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
\author{John R. Tipton \orcidaffilb{}}
%\affil{Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, USA}
\author{Timothy J. Muldoon*}
%\affil{Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, USA}
\affil{tmuldoon@uark.edu}

@@

%\lstset{
%  breaklines=true,
%  stringstyle=\ttfamily,
%  backgroundcolor=\color{gray}
%}
\usepackage{placeins}
\usepackage{subfig}
\usepackage{breqn}
\usepackage[font={small}]{caption}
\usepackage{float}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{SUPPLEMENTARY MATERIALS for\\
\strut \\
\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\
Beyond repeated measures ANOVA and Linear Mixed Models\\
\strut \\
APPENDIX A: WORKFLOW FOR GAMs FOR BIOMEDICAL LONGITUDINAL DATA}
\author{}
\date{\vspace{-2.5em}}

@@
